Modularly Composing Typed Language Fragments

### 2014-11-01 att-pldi15/att-pldi15.tex
More revisions

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show ordered record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of associated term-level operators using functions written in a static language where types and translations are values. The semantics maintain several strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon  invariants} maintained under a minimal tycon context will be conserved under any further extended context. Interestingly, mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies reasoning about ML-style modules.
\end{abstract}

@@

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. \verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons in Fig. \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

@@

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\\\
    \psi = (\tcsig{\ktyidx}{\chi})\\
    \keq{\emptyset}{\ktyidx}\\\\
    \theta = (\tcstruct{\stx{schema}}{\omega})\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi, \tcdef{\tc}{\psi}{\theta}} \omega \sim \psi
}{
    \vdash \Phi, \tcdef{\tc}{\psi}{\theta}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the empty case (tcc-emp) for concision.}
\label{fig:tycon-ctxs}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\subsection{Typing Contexts}
We must also define a translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ that checks that $\Upsilon$ contains valid types, then generates the corresponding internal typing context $\Gamma$ from their translations (see supplement).
%\vspace{-5px}
 %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
\section{Typing and Translation of External Terms}\label{external-terms}
Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.

@@

\inferrule[ocstruct-targ]{
    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
    \opname{op}\notin\text{dom}(\chi)\\
    \emptyset \vdash \klitidx\\\\
    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
}{
    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
}\vspace{-5px}
\end{mathpar}
\caption{Checking opcon structures against tycon signatures.}
\label{ocstruct}
\end{figure*}

Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where we define an example typing context $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$. The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).

@@

The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$. This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so. The function has access to the type index, the term index and an interface to the list of arguments. The  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext),  checks that the intro opcon definition is well-kinded: rule (ocstruct-intro) in Figure \ref{ocstruct}.

For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
\[\small\begin{array}{l}
    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\\
\quad \svar{str\_of\_Str}~\svar{tmidx}
}}}
\end{array}\]
Because regular strings are implemented as strings, this intro opcon definition is straightforward. It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$. In practice, the tycon provider would specify an error message here.%The kinding rules and operational semantics of this exception-like form are standard:

@@

The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}. This form is analagous to the introductory form for kind $\kity$ described previously, $\sqity{\qity}$, where $\qity$ is a translational internal type. Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and operational semantics simply recurse through these in the same manner as shown in Sec. \ref{sec:type-translations}. Also analagously, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally. The supplement gives the analagous rules.

The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$. These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis. They are not intended to be written directly by tycon providers, arising only internally in the semantics of argument interface lists. Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$. We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
\[\small
\begin{array}{l}
\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
}
\end{array}
\]
The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are. An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML. We chose not to formalize these for simplicity, and to demonstrate this general technique. An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited. In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.

The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces. We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length. The base case is the translational empty product. The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$. Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$, which  produces a translational internal term  referring to the translation of that argument if it succeeds, discussed next. The final line constructs a nested tuple based on this translation and the recursive result. Taken together, the translational internal term that will be derived for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.

@@

The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$ (see supplement).

@@

\paragraph{Abstracted Term Translations} The reason that the translations themselves were not inserted yet (even though they were derived in rule (s-ana-success) above) is again because we want to be able to  hold these abstract by replacing them with variables. The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec. \ref{sec:selective-type-translation-abstraction}. The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings. Term translation stores have the following syntax: $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$. Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''. For example, \[\small\edeabs{\tcvar{lprod}}{(title, \texttt{"EXMPL 2015"}),\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec. \ref{sec:selective-type-translation-abstraction}.

The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement). The rules for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
\begin{mathpar}\small
\inferrule[abs-anatrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \eanaX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{\tc}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
}
\end{mathpar}

@@

The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon. The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation. In all other respects, the rule is analagous. It is used as a technical device in our proof of conservativity in Sec. \ref{metatheory}.

\subsection{Generalized Targeted Operations} 
All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments. Some examples of desugarings to this form include explicit invocation, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$. We show other desugarings, including case analysis, in the supplement.

@@

\begin{theorem}[Static Preservation]
If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
\end{theorem}

The case for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined. The mutual induction is well-founded because the total length of all the argument lists in the terms being considered monotonically decreases.
\begin{theorem}[Type Synthesis]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
\end{theorem}
\paragraph{Type Safety}
Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms. Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong. To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}).% produced as the translation of the external type. Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.

>>>>

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of associated term-level operators (e.g. record projection) using functions written in a static language where types and translations are values. The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension. Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
\end{abstract}

@@

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

@@

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\subsection{Typing Contexts}
We must also define a translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ that checks that $\Upsilon$ contains valid types, then generates the corresponding internal typing context $\Gamma$ from their translations (see supplement).
%\vspace{-5px}
 %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
\section{Typing and Translation of External Terms}\label{external-terms}

Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.

@@

\inferrule[ocstruct-targ]{
    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
    \opname{op}\notin\text{dom}(\chi)\\
    \emptyset \vdash \klitidx\\\\
    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
}{
    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
}\vspace{-5px}
\end{mathpar}
\caption{Checking opcon structures against tycon signatures.}
\label{ocstruct}
\end{figure*}
Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where we define an example typing context $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$. The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).

@@

The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$. This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so. The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.

For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
\[\small\begin{array}{l}
    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
}}}
\end{array}\]
Because regular strings are implemented as strings, this intro opcon definition is straightforward. It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$. In practice, the tycon provider would specify an error message here.%The kinding rules and operational semantics of this exception-like form are standard:

@@

\paragraph{Term Translations} The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}. This form is analagous to the introductory form for kind $\kity$ described in Sec. \ref{sec:type-translations}, $\sqity{\qity}$. Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and operational semantics simply recurse through these in the same manner as shown in Sec. \ref{sec:type-translations}. Also analagously, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally. The supplement gives the analagous rules.

The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$. These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis. They are not intended to be written directly by tycon providers, arising only internally via argument interface lists. Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$. We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
\[\small
\begin{array}{l}
\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
}
\end{array}
\]
The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are. An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML. We chose not to formalize these for simplicity, and to demonstrate this general technique. An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited. In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.

The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces. We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length. The base case is the translational empty product. The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$. Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$. Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$ where $\st$ is the value of $\svar{rowty}$ if the row value is well-typed and fails if not. The final line constructs a nested tuple based on this translation and the recursive result. Taken together, the translational internal term that will be derived for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.

@@

The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$ (see supplement).

@@

\paragraph{Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables. The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec. \ref{sec:selective-type-translation-abstraction}. The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings. Term translation stores have the following syntax: $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$. Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''. For example, \[\small\edeabs{\tcvar{lprod}}{(title, \texttt{"EXMPL 2015"}),\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec. \ref{sec:selective-type-translation-abstraction}.

The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement). The rules for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
\begin{mathpar}\small
\inferrule[abs-anatrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \eanaX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{c}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
}
\end{mathpar}

@@

\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon. The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation. In all other respects, the rule is analagous. It is used as a technical device in our proof of conservativity in Sec. \ref{metatheory}.

\subsection{Generalized Targeted Operations} 
All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.

Desugarings to this form include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$. We show other desugarings, including case analysis, in the supplement.

@@

\begin{theorem}[Static Preservation]
If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
\end{theorem}
The case in the proof for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined. The mutual induction is well-founded because the total length of all the argument lists in the terms being considered decreases.
\begin{theorem}[Type Synthesis]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
\end{theorem}
\paragraph{Type Safety}
Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms. Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong. To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}):% produced as the translation of the external type. Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
